\emph{Predictive modeling in the financial sector is severely hindered by class imbalance, where critical events such as fraud, loan defaults, and bankruptcies are exceedingly rare, constituting less than 5\% of observations in most real-world datasets. Standard generative methods, including CTGAN, TVAE, and TabSyn, faithfully replicate these skewed distributions and offer no mechanism to adjust the class ratio at generation time, while direct diffusion approaches such as TabDDPM fail on mixed-type tabular data due to discontinuous gradients introduced by one-hot encoded categorical features (KS $>$ 0.80). This work presents RE-TabSyn (Rare-Event Enhanced Tabular Synthesis), a framework that applies Classifier-Free Guidance (CFG) with latent diffusion models to enable explicit, inference-time control over synthetic class proportions in tabular data. RE-TabSyn is built on three components: a Variational Autoencoder that encodes heterogeneous tabular features, combining numerical and categorical attributes, into a smooth continuous latent space; a Transformer-based Latent Diffusion Model that generates high-fidelity samples within that latent space; and a Classifier-Free Guidance mechanism that allows practitioners to specify target minority ratios through a single guidance scale parameter $w$, without retraining the model. To the best of the authors' knowledge, this is the first application of CFG to tabular data synthesis, confirmed through a systematic review of 151 research papers across 10 categories. RE-TabSyn is evaluated on six financial benchmark datasets covering credit risk, income prediction, marketing response, loan performance, and corporate bankruptcy, with minority ratios ranging from 4.8\% to 44.5\%. Experiments conducted across three random seeds establish three principal findings. First, RE-TabSyn reliably generates balanced datasets, achieving approximately 50\% minority representation within 1.7\% of the target on average, boosting minority representation by up to $10\times$ compared to the original training data. Second, classifiers trained on RE-TabSyn's balanced synthetic data achieve a mean minority F1-score of 0.472, surpassing the real imbalanced data baseline of 0.458, despite a controlled fidelity trade-off (KS: 0.109 for TabSyn versus 0.171 for RE-TabSyn). Third, no systematic memorization of training data is observed, with Distance to Closest Record exceeding 1.0 in mean across all datasets. The core findings have been accepted for presentation at the I2IT International Conference, confirming the scientific merit and novelty of the proposed approach.}
